---
title: "Exercise 3 - Classification"
author: "Ken Benoit"
output: html_document
---

You will need to load the core library for the course textbook:
```{r}
library(ISLR)
```


## Exercise 3.1

This question should be answered using the `Weekly` data set, which is part of the `ISLR` package. This data contains 1,089 weekly stock returns for 21 years, from the beginning of 1990 to the end of 2010.

(a) Produce some numerical and graphical summaries of the `Weekly` data. Do there appear to be any patterns?

```{r}
library(ISLR)
summary(Weekly)
pairs(Weekly)
cor(Weekly[,-9])
```

**Year and Volume appear to have a relationship. No other patterns are discernible.**

(b) Use the full data set to perform a logistic regression with `Direction` as the response and the five lag variables plus `Volume` as predictors. Use the summary function to print the results. Do any of the predictors appear to be statistically significant? If so, which ones?

```{r}
attach(Weekly)
glm.fit <-  glm(Direction ~ Lag1 + Lag2 + Lag3 + Lag4 + Lag5 + Volume,
              data = Weekly,
              family = binomial)
summary(glm.fit)
```

**Lag 2 appears to have some statistical significance with a Pr(>|z|) = 3%.**

(c) Compute the confusion matrix and overall fraction of correct predictions. Explain what the confusion matrix is telling you about the types of mistakes made by logistic regression.

```{r}
glm.probs <-  predict(glm.fit, type="response")
glm.pred <-  rep("Down", length(glm.probs))
glm.pred[glm.probs>.5] <-  "Up"
table(glm.pred, Direction)
```

**Percentage of currect predictions: (54+557)/(54+557+48+430) = 56.1%. Weeks the market goes up the logistic regression is right most of the time, 557/(557+48) = 92.1%. Weeks the market goes up the logistic regression is wrong most of the time 54/(430+54) = 11.2%.**

(d) Now fit the logistic regression model using a training data period from 1990 to 2008, with `Lag2` as the only predictor. Compute the confusion matrix and the overall fraction of correct predictions for the held out data (that is, the data from 2009 and 2010).

```{r}
train <-  (Year < 2009)
Weekly.0910 <-  Weekly[!train,]
glm.fit <-  glm(Direction ~ Lag2, data = Weekly, family = binomial, subset = train)
glm.probs <-  predict(glm.fit, Weekly.0910, type="response")
glm.pred <-  rep("Down", length(glm.probs))
glm.pred[glm.probs>.5] <-  "Up"
Direction.0910 <-  Direction[!train]
table(glm.pred, Direction.0910)
mean(glm.pred == Direction.0910)
```


(e) Repeat (d) using LDA.

```{r}
library(MASS)
lda.fit <-  lda(Direction ~ Lag2, data = Weekly, subset = train)
lda.pred <-  predict(lda.fit, Weekly.0910)
table(lda.pred$class, Direction.0910)
mean(lda.pred$class == Direction.0910)
```

(f) Repeat (d) using QDA.

```{r}
qda.fit <-  qda(Direction ~ Lag2, data = Weekly, subset = train)
qda.class <-  predict(qda.fit, Weekly.0910)$class
table(qda.class, Direction.0910)
mean(qda.class == Direction.0910)
```

**A correctness of 58.7% even though it picked Up the whole time.**

(g) Repeat (d) using KNN with K = 1.

```{r}
library(class)
train.X <-  as.matrix(Lag2[train])
test.X <-  as.matrix(Lag2[!train])
train.Direction <-  Direction[train]
set.seed(1)
knn.pred <-  knn(train.X, test.X, train.Direction, k = 1)
table(knn.pred, Direction.0910)
mean(knn.pred == Direction.0910)
```

(h) Which of these methods appears to provide the best results on this data?

**Logistic regression and LDA methods provide similar test error rates.**

(i) Experiment with different combinations of predictors, including possible transformations and interactions, for each of the methods. Report the variables, method, and associated confusion matrix that appears to provide the best results on the held out data. Note that you should also experiment with values for K in the KNN classifier.

```{r}
# Logistic regression with Lag2:Lag1
glm.fit <-  glm(Direction~Lag2:Lag1, data = Weekly, family = binomial, subset = train)
glm.probs <-  predict(glm.fit, Weekly.0910, type = "response")
glm.pred <-  rep("Down", length(glm.probs))
glm.pred[glm.probs>.5] <-  "Up"
Direction.0910 <-  Direction[!train]
table(glm.pred, Direction.0910)
mean(glm.pred == Direction.0910)

# LDA with Lag2 interaction with Lag1
lda.fit <-  lda(Direction ~ Lag2:Lag1, data = Weekly, subset = train)
lda.pred <-  predict(lda.fit, Weekly.0910)
mean(lda.pred$class == Direction.0910)

# QDA with sqrt(abs(Lag2))
qda.fit <-  qda(Direction ~ Lag2 + sqrt(abs(Lag2)), data = Weekly, subset = train)
qda.class <-  predict(qda.fit, Weekly.0910)$class
table(qda.class, Direction.0910)
mean(qda.class == Direction.0910)

# KNN k =10
knn.pred <-  knn(train.X, test.X, train.Direction, k = 10)
table(knn.pred, Direction.0910)
mean(knn.pred == Direction.0910)

# KNN k = 100
knn.pred <-  knn(train.X, test.X, train.Direction, k = 100)
table(knn.pred, Direction.0910)
mean(knn.pred == Direction.0910)
```

**Out of these permutations, the original LDA and logistic regression have better performance in terms of test error rate.**


## Exercise 3.2

In this problem, you will develop a model to predict whether a given car gets high or low gas mileage based on the `Auto` dataset from the `ISLR` package.

(a) Create a binary variable, `mpg01`, that contains a 1 if `mpg` contains a value above its median, and a 0 if `mpg` contains a value below its median. You can compute the median using the `median()` function. Note you may find it helpful to use the `data.frame()` function to create a single data set containing both `mpg01` and the other `Auto` variables.

```{r}
library(ISLR)
summary(Auto)
attach(Auto)
mpg01 <-  rep(0, length(mpg))
mpg01[mpg>median(mpg)] <- 1
Auto <- data.frame(Auto, mpg01)
```


(b) Explore the data graphically in order to investigate the association between `mpg01` and the other features. Which of the other features seem most likely to be useful in predicting `mpg01`? Scatterplots and boxplots may be useful tools to answer this question. Describe your findings.

```{r 11b}
cor(Auto[,-9])
pairs(Auto) # doesn't work well since mpg01 is 0 or 1
```

**cylinders, weight, displacement, horsepower (and mpg itself)**

(c) Split the data into a training set and a test set.

```{r}
train <-  (year %% 2 == 0) # if the year is even
test <-  !train
Auto.train <-  Auto[train,]
Auto.test <-  Auto[test,]
mpg01.test <-  mpg01[test]
```

(d) Perform LDA on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?

```{r}
# LDA
library(MASS)
lda.fit <-  lda(mpg01 ~ cylinders + weight + displacement + horsepower,
              data = Auto, subset = train)
lda.pred <-  predict(lda.fit, Auto.test)
mean(lda.pred$class != mpg01.test)
```

**12.6% test error rate.**

(e) Perform QDA on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?

```{r}
# QDA
qda.fit <-  qda(mpg01 ~ cylinders + weight + displacement + horsepower,
              data = Auto, subset = train)
qda.pred <-  predict(qda.fit, Auto.test)
mean(qda.pred$class != mpg01.test)
```

**13.2% test error rate.**


(f) Perform logistic regression on the training data in order to predict `mpg01` using the variables that seemed most associated with `mpg01` in (b). What is the test error of the model obtained?

```{r}
# Logistic regression
glm.fit <-  glm(mpg01 ~ cylinders + weight + displacement + horsepower,
              data = Auto,
              family = binomial,
              subset = train)
glm.probs <-  predict(glm.fit, Auto.test, type = "response")
glm.pred <-  rep(0, length(glm.probs))
glm.pred[glm.probs > 0.5] <- 1
mean(glm.pred != mpg01.test)
```

**12.1% test error rate.**

(g) Perform KNN on the training data, with several values of K, in order to predict `mpg01`. Use only the variables that seemed most associated with `mpg01` in (b). What test errors do you obtain? Which value of K seems to perform the best on this data set?

```{r}
library(class)
train.X <-  cbind(cylinders, weight, displacement, horsepower)[train,]
test.X <-  cbind(cylinders, weight, displacement, horsepower)[test,]
train.mpg01 <-  mpg01[train]
set.seed(1)

# KNN (k=1)
knn.pred <-  knn(train.X, test.X, train.mpg01, k = 1)
mean(knn.pred != mpg01.test)

# KNN (k=10)
knn.pred <-  knn(train.X, test.X, train.mpg01, k = 10)
mean(knn.pred != mpg01.test)

# KNN (k=100)
knn.pred <-  knn(train.X, test.X, train.mpg01, k = 100)
mean(knn.pred != mpg01.test)
```

**k=1, 15.4% test error rate. k=10, 16.5% test error rate. k=100, 14.3% test error rate. K of 100 seems to perform the best. 100 nearest neighbors.**


## Exercise 3.3 (Optional)

This problem involves writing functions.

(a) Write a function, `Power()`, that prints out the result of raising 2 to the 3rd power. In other words, your function should compute $2^3$ and print out the results.

    *Hint: Recall that `x^a` raises `x` to the power `a`. Use the `print()` function to output the result.*


```{r}
Power <-  function() {
  2^3
}
print(Power())
```


(b) Create a new function, `Power2()`, that allows you to pass *any* two numbers, `x` and `a`, and prints out the value of `x^a`. You can do this by beginning your function with the line
```{r eval=FALSE}
Power2 <- function(x,a) {
```
You should be able to call your function by entering, for instance,
```{r eval=FALSE}
Power2(3,8)
```
on the command line. This should output the value of $3^8$, namely, 6,561.


```{r}
Power2 <-  function(x, a) {
  x^a
}
Power2(3,8)
```




(c) Using the `Power2()` function that you just wrote, compute $10^3$, $8^{17}$, and $131^3$.

```{r}
Power2(10, 3)
Power2(8, 17)
Power2(131, 3)
```


(d) Now create a new function, `Power3()`, that actually returns the result `x^a` as an R object, rather than simply printing it to the screen. That is, if you store the value `x^a` in an object called `result` within your function, then you can simply `return()` this result, using the following line:

```{r eval=FALSE}
return(result)
```

The line above should be the last line in your function, before the `}` symbol.

```{r}
Power3 <-  function(x, a) {
  result <-  x^a
  return(result)
}
```


(e) Now using the `Power3()` function, create a plot of $f(x) = x^2$. The $x$-axis should display a range of integers from 1 to 10, and the $y$-axis should display $x^2$. Label the axes appropriately, and use an appropriate title for the figure. Consider displaying either the $x$-axis, the $y$-axis, or both on the log-scale. You can do this by using `log="x"`, `log="y"`, or `log="xy"` as arguments to the `plot()` function.

```{r}
x <-  1:10
plot(x, Power3(x, 2),  
     log="xy", ylab="Log of y = x^2", xlab="Log of x", 
     main="Log of x^2 versus Log of x")
```


(f) Create a function, `PlotPower()`, that allows you to create a plot of `x` against `x^a` for a fixed `a` and for a range of values of `x`. For instance, if you call
```{r eval=FALSE}
PlotPower(1:10, 3)
```
then a plot should be created with an $x$-axis taking on values $1,2,\dots ,10$, and a $y$-axis taking on values $1^3,2^3,\dots ,10^3$.

```{r}
PlotPower <-  function(x, a) {
  plot(x, Power3(x, a))
}
PlotPower(1:10, 3)
```

